The interactions of rational, pragmatic agents  \n lead to efficient language structure and use

### 2019-01-31 paper/paper.Rmd
updates from MF / BP edits

abstract: 
    Languages display a diverse set of distributional regularities such as the relation between a word's frequency and rank in a corpus, the distribution of syntactic dependency lengths across languages, or the presence of properties such as ambiguity. We discuss a framework for studying how these system-level properties emerge from local, in-the-moment interactions of rational, pragmatic speakers and listeners. To do so, we derive a novel objective function for measuring the communicative efficiency of linguistic systems in terms of the interactions of speakers and listeners. We examine the behavior of this objective in a series of simulations focusing on the communicative function of ambiguity in language. These simulations suggest that rational pragmatic agents will produce communicatively efficient systems and that interactions between such agents provide a framework for examining efficient properties of language structure and use more broadly.

@@

But how does this optimization take place? The example given by Zipf [-@Zipf1949a], describes local, communicative interactions in terms of a \textit{reference game}. Speakers intend to refer to some object in the world $m$. They choose some utterance $u$ to transmit this intended meaning, $u \rightarrow m$. The listener attempts to reconstruct this intended meaning given the transmitted utterance, $m \rightarrow u$. Other projects have assumed this basic reference game setting [@RegierKempKay2015a; @Piantadosi2011a] and this simplification of the communicative act has proven productive in theoretical [@FerreriCancho2018a], simulation-based [@KirbyGriffithsSmith2014a] and empirical explorations [@HawkinsFrankeSmithGoodman2018a] of efficient language structure and use.\par

Zipf's conception of speaker and listener effort may be connected to speakers' pragmatic reasoning in the moment. Under a Gricean treatment of pragmatics, speakers and listeners follow a set of conversational maxims in which they cooperate to transfer information [@Grice1975a]. These maxims appear to emerge from efficiency concerns, however (Horn, 1984). We formalize this connection -- showing how system-level efficiencies can emerge from local interaction behavior. Intuitively, our claim is that to understand an "efficient" property of a system it is essential that we consider how that property is *used* efficiently. It follows that any quantitative measure of "efficiency" must be a function of both the system and its users.\par

In the current work we provide a case-study for the framing presented above. We choose a property of languages that could, in principle, vary freely, but shows strong regularities across languages. The explanadum is why this regularity holds. We examine ambiguity as our property, extending ideas by Piantadosi et al. [-@Piantadosi2011a]. We define a novel measure of efficiency that depends on the interactional behavior of speaker and listener agents. We adopt the reference game as our primary unit of interaction and model language users with the Rational Speech Act (RSA) framework -- a computional model of language use, which is nicely attested by experimental data on interaction. Using these basic ingredients, we show that the property of interest (ambiguity) is prevalent in languages that optimize our measure of efficiency. Further, we show how ambiguity is \textit{used} efficiently during local, in-the-moment interactions. The contributions of this work are twofold -- we derive a novel objective for measuring linguistic efficiency and also show how the reference game framework, in combination with formal models of communication, can be used to connect ideas about system-level efficiencies to in-the-moment language use.\par

@@

### Speakers and listeners
The Rational Speech Act framework (RSA) is a computational-level theory of pragmatic language use, which has been well attested in experimental work examining both speaker and listener communication behavior [@FrankGoodman2012a; @GoodmanFrank2016a]. RSA is a formalization of essential Gricean pragmatic principles -- agents reason about one another and their shared context [@Grice1975a]. We adopt RSA as our representational framework to model Gricean (rational and pragmatic) speaker-listeners in the reference game setting\footnote{See SI section 1 for a more detailed description of basic RSA speaker-listener defintions and to Goodman \& Frank (2016) for an overview of its applications.}.\par

@@

We assume that speaker effort is related to the surprisal of an utterance in a particular context\footnote{In the current set of simulations we consider utterances costs as independent from context (ie. $p(u|c)p(c)=p(u)p(c)$).} -- intuitively, the number of bits needed to encode the utterance $u$. This particular formalization of speaker-cost is general enough to accommodate a range of cost instantiations, such as production difficult via articulation effort, cognitive effort related to lexical access, or others [@BennettGoodman2015a].\par

@@

From an information-theoretic perspective this objective is intuitive: The Cross-Entropy (CE) gives us a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, we have an expectation over this term -- the expected difference between the distributions assumed by the speaker $P_{speaker}$ and listener $P_{listener}$ given a set of contexts $C$\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy over utterance-meaning pairs $H_{cross}[P_{speaker}(u, m; \ell), P_{listener}(u, m; \ell)]$}. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listeners think.

@@

In the following experiments we explore two aspects of Piantadosi’s claim. In Simulation 1, we examine the efficient language *structure* aspect of their claim, exploring when the optimal linguistic system $\ell^*$ is most likely to contain ambiguous lexical items. In Simulation 2, we explore an efficient language *use* aspect of the claim -- at what point in a conversation is it useful for a speaker to use ambiguous lexical material?\par

@@

We adopt the hypothesis presented by Piantadosi, formalizing their argument in our current framework -- showing that ambiguity is an efficient property under our CE objective in the reference game setting. We proceed by generating languages with different amounts of contextual support (varying the size of $|C|$). We search the space of languages, examining whether ones which minimize our objective contain ambiguity. If context leads to more efficient communication, then optimal languages under our objective should be more likley to be ambiguous as the amount of context increases.\par

@@

Following the argument given by Piantadosi et al. [-@Piantadosi2011a], we want to assess the impact of *context* on the presence of ambiguity in optimal languages. To do so we consider four conditions with $n=500$ simulations each (that is, 500 unique sets of $\{P(U), P(M|C), P(C)\}$. Our first is a \textit{one-context} condition ($|C|=1$) -- only a single distribution over meanings $P(M)$. In our \textit{two-context} condition ($|C| = 2$), we consider efficiency under both $P(M|C=c_1)$ as well as $P(M|C=c_2)$. \textit{Three-} and \textit{four-context} conditions corresponding accordingly.\par

@@

In Simulation 1 we explored the degree to which ambiguity is an efficient property of languages when communication is contextualized. Figure 1. plots the proportion of optimal languages under each objective as a function of context size. The red line shows that as the number of contexts increases, so does the probability that an optimal language $\ell^*_{cross}$ contains ambiguity under our CE objective. For comparison we also plot the proportion of speaker-optimal $\ell^*_{speaker}$ (blue line) and listener-optimal $\ell^*_{listener}$ (green line) languages that contain ambiguity. In line with Zipf's predictions, if languages are designed only to minimize speaker effort then optimal languages always contain ambiguity. If languages are designed to minimize listener effort then ambiguity is always avoided.\par

While our results indicate that ambiguity is an efficient property of contextualized language, these simulations assumed that agents could always disambiguate items in context. That is, in our four conditions both agents had perfect knowledge of the relevant conditional distributions ($P(M|C)$). This assumption may be too strong for describing much of day-to-day communication -- we seldom interact with others with perfect knowledge of the current context (or topic) at the start of a conversation. To explore how ambiguity may be \textit{used} efficiently in our framework, we next examine a case in which the listener has imperfect knowledge of context at the start of the conversation, but may infer it from the contents of discourse.\par

@@

We shift focus in Simulation 2 -- instead of examining the landscape of a space of languages under our CE objective, we consider a single language containing ambiguity, examining how that property is used over a discourse. If context is not informative with respect to meaning early in a discourse, but is at later positions, then an efficient speaking strategy would avoid ambiguous material early, saving it for when contextual information is disambiguating. We can consider “context” as analogous to a “topic” of conversation. As an example, imagine a scenario in which a reader is beginning a magazine article. While they may have some knowledge about the article’s topic (perhaps from the title), they may not have complete knowledge of its contents, including the persons or events involved. In this setting, using a low-cost, but ambiguous pronoun early may lead to misunderstanding if context is not informative. But, if by a later position enough contextual information has accumulated, it may be efficient to use the ambiguous material.\par

@@

We consider consider a single language $\ell$, which contains both ambiguous and unambiguous utterances. We assume ambiguous utterances are lower cost. Crucially, we do not assume that the listener knows the particular topic ($c_{current}$) of the conversation \textit{a priori}. Rather, that the listener has knowledge of the set of possible topics $C = \{c_1, \dots, c_k\}$, but \textit{does not know which one is currently being used by the speaker}. Formally, this means the listener does not have access to the correct conditional distribution over meanings $P(M|C=c_{current})$ at the start of the discourse.\par

@@

We conduct $N=600$ simulations, generating discourses of length $|D|=30$ utterances, comparing three speaker models ($n=200$ each). We consider a single language $\ell$\footnote{See SI for the matrix notation of this langauge.} with $|U|=6$ and $|M|=4$ in which two of the utterances are ambiguous and lower cost than the unambiguous utterances. (Note that use of this particular language is not essential -- the results are broadly generalizable to languages that contain ambiguity.)\par

## Speaker agents
We consider three types of speaker models. Our \textit{Full pragmatics} agent, models a speaker who reasons about her listener and also has complete recall of the set of utterances in the discourse $D$. This speaker believes that the listener may not know the current topic $c_{current}$ at the start of the discrouse, but can infer it over the discourse. We compare two baseline models. The first, a \textit{Partial pragmatics} baseline describes a speaker who reasons about a listener, but assumes they have no access to discourse history. The second, a \textit{No pragmatics} speaker who does not consider a listener at all, but produces utterances according to the underlying langauge semantics ($\ell$) and topic probabilities ($p(M|C=c_{current}$) (see SI section 3 for details).\par

@@

We are interested in \textit{when} it is efficient to use low-cost, ambiguous material in this discourse setting. A speaker strategy that is mutually efficient for both agents should avoid ambiguity until sufficient contextual information has accumulated. We should expect this to be reflected in our \textit{Full pragmatics} model who reasons about the listener and discourse history. By contrast, a speaker-optimal model who does not consider the listener should greedily use ambiguous utterances (\textit{No Pragmatics} model), while a listener-optimal model should avoid ambiguity entirely (\textit{Partial pragmatics} model).\par

@@

Figure 2, (A) shows the empirical probability that a speaker uses an ambiguous utterance as a function of discourse position. The \textit{No pragmatic} baseline uses ambiguous utterances frequently and at a constant rate over the discourse and the \textit{Partial pragmatic} baseline avoids ambiguous utterances entirely. But, the \textit{Full pragmatic} model avoids ambiguous material only at the start of the discourse, employing it increasingly as the discourse proceeds. (C) tracks our CE objective for each model over the discourse. Note that the objective decreases for all three models, primarily driven by the listener updating his belief about the actual topic ($P(C=c_{current}|D)$). However, the objective declines more quickly under the \textit{Full} and \textit{Partial} pragmatic speakers as listener agents are better able to infer the correct context. Additionally, the difference in CE between the \textit{Full}- and \textit{Partial pragmatic} models at the end of the discourse is driven by the reduction in speaker costs described in (B). While speaker effort remains constant in both \textit{No pragmatic} and \textit{Partial pragmatic} baselines, effort declines in the \textit{Full pragmatic} model as she increasingly relies on ambiguous material later in the discourse.\par

@@

How do the competing pressures of speakers and listeners give rise to the distributional forms found in natural language? Zipf [-@Zipf1949a] proposed that the asymmetry between speaker and listener costs gives rise to a range of properties at the level of the lexicon. We explored the interactions of rational pragmatic agents as a framework for understanding efficient language structure and use. We focused on an argument on the communicative function of ambiguity [@Piantadosi2011a], deriving a novel speaker-listener Cross-Entropy objective for measuring the efficiency linguistic systems from first principles of efficient language use. In a series of simulations we showed that optimal languages are more likely to contain ambiguous material when context is informative and that rational pragmatic agents will *use* ambiguous material efficiently.\par

A limitation of the current work is an analysis of exactly how this objective compares to existing measures. For example, previous work has described competing speaker-listener pressures in terms of a trade-off of \textit{simplicity} and \textit{informativeness} [@KempRegier2012a] or \textit{expressivity} and \textit{compressibility} [@SmithTamarizKirby2013a]. Future work should assess the degree to which we can derive the same properties as previous studies using our current framework. In line with this idea, a primary focus of the current work is to provide a case-study in the productivity of assessing functionalist theories in repeated reference games between rational pragmatic agents. In theory, this framework can serve as a domain general tool to assess the range of functionalist theories examining efficient language-structure and use.\par

>>>>

abstract: 
    Despite their diversity, languages around the world share a consistent set of properties and distributional regularities. For example, the distribution of word frequencies, the distribution of syntactic dependency lengths, and the presence of ambiguity are all remarkably consistent across languages. We discuss a framework for studying how these system-level properties emerge from local, in-the-moment interactions of rational, pragmatic speakers and listeners. To do so, we derive a novel objective function for measuring the communicative efficiency of linguistic systems in terms of the interactions of speakers and listeners. We examine the behavior of this objective in a series of simulations focusing on the communicative function of ambiguity in language. These simulations suggest that rational pragmatic agents will produce communicatively efficient systems and that interactions between such agents provide a framework for examining efficient properties of language structure and use more broadly.

@@

But how does this optimization take place? The example given by Zipf [-@Zipf1949a] describes local, communicative interactions in terms of a \textit{reference game}. Speakers intend to refer to some object in the world $m$. They choose some utterance $u$ to transmit this intended meaning, $u \rightarrow m$. The listener attempts to reconstruct this intended meaning given the transmitted utterance, $m \rightarrow u$. Other projects have assumed this basic reference game setting [@RegierKempKay2015a; @Piantadosi2011a] and this simplification of the communicative act has proven productive in theoretical [@FerreriCancho2018a], simulation-based [@KirbyGriffithsSmith2014a] and empirical explorations [@HawkinsFrankeSmithGoodman2018a] of efficient language structure and use.\par

Adopting reference games as a basic unit of analysis suggests that optimization may take place at the level of conversation. Importantly, Zipf's conception of speaker and listener effort may be connected to how language is used; in particular, whether interlocutors engage in pragmatic reasoning during conversation. Under a Gricean treatment of pragmatics, speakers and listeners follow a set of conversational maxims in which they cooperate to transfer information [@Grice1975a]. These maxims appear to emerge from efficiency concerns, however (Horn, 1984). We formalize this connection -- showing how system-level efficiencies can emerge from local interaction behavior of pragmatic agents. Intuitively, our claim is that to understand an "efficient" property of a system it is essential that we consider how that property is *used* efficiently.\par

We provide a case study for this approach, in which functionalist regularities emerge from the dynamics of pragmatic communication. We choose a property of languages that could, in principle, vary freely, but shows strong regularities across languages. The explanadum is why this regularity holds. We examine ambiguity as our property, extending ideas by Piantadosi et al. [-@Piantadosi2011a]. We define a novel measure of efficiency that depends on the interactional behavior of speaker and listener agents. We adopt the reference game as our primary unit of interaction and model language users with the Rational Speech Act (RSA) framework -- a computional model of language use, which is nicely attested by experimental data on interaction. Using these basic ingredients, we show that the property of interest (ambiguity) is prevalent in languages that optimize our measure of efficiency. Further, we show how ambiguity is \textit{used} efficiently during local, in-the-moment interactions. The contributions of this work are twofold -- we derive a novel objective for measuring linguistic efficiency and also show how the reference game framework, in combination with formal models of communication, can be used to connect ideas about system-level efficiencies to in-the-moment language use.\par

@@

### Speakers and listeners
The Rational Speech Act framework (RSA) is a computational-level theory of pragmatic language use, which has produced good fit to human communication behavior across a range of language phenomena [@FrankGoodman2012a; @GoodmanFrank2016a]. RSA is a formalization of essential Gricean pragmatic principles -- agents reason about one another and their shared context [@Grice1975a]. We adopt RSA as our representational framework to model Gricean (rational and pragmatic) speaker-listeners in the reference game setting (see SI and @GoodmanFrank2016a for review).\par

@@

We assume that speaker effort is related to the surprisal of an utterance in a particular context\footnote{In the current set of simulations we consider utterances costs as independent from context (ie. $p(u|c)p(c)=p(u)p(c)$).} -- intuitively, the number of bits needed to encode the utterance $u$. This particular formalization of speaker-cost is general enough to accommodate a range of cost instantiations, such as production difficulty via articulation effort, cognitive effort related to lexical access, or others [@BennettGoodman2015a].\par

@@

From an information-theoretic perspective this objective is intuitive: $H_{cross}$ denotes the Cross-Entropy (CE), a measure of dissimilarity between two distributions -- the average number of bits required to communicate under one distribution, given that the “true” distribution differs. In our case, we have an expectation over this term -- the expected difference between the distributions assumed by the speaker $P_{speaker}$ and listener $P_{listener}$ given a set of contexts $C$\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy over utterance-meaning pairs $H_{cross}[P_{speaker}(u, m; \ell), P_{listener}(u, m; \ell)]$}. In other words, an "efficient" language $\ell$ minimizes the distance between what speakers and listeners think.

@@

In the following experiments we explore two aspects of Piantadosi’s et al.'s claim. In Simulation 1, we examine the efficient language *structure* aspect of their claim, exploring when the optimal linguistic system $\ell^*$ is most likely to contain ambiguous lexical items. In Simulation 2, we explore an efficient language *use* aspect of the claim -- at what point in a conversation is it useful for a speaker to use ambiguous lexical material?\par

@@

We show that ambiguity is an efficient property under our CE objective in the reference game setting. We proceed by generating languages with different amounts of contextual support (varying the size of $|C|$). We search the space of languages, examining whether ones which minimize our objective contain ambiguity. If context leads to more efficient communication, then optimal languages under our objective should be more likely to be ambiguous as the amount of context increases.\par

@@

We want to assess the impact of *context* on the presence of ambiguity in optimal languages. To do so we consider four conditions with $n=500$ simulations each (that is, 500 unique sets of $\{P(U), P(M|C), P(C)\}$. Our first is a \textit{one-context} condition ($|C|=1$) -- only a single distribution over meanings $P(M)$. In our \textit{two-context} condition ($|C| = 2$), we consider efficiency under both $P(M|C=c_1)$ as well as $P(M|C=c_2)$. \textit{Three-} and \textit{four-context} conditions corresponding accordingly.\par

@@

In Simulation 1 we explored the degree to which ambiguity is an efficient property of languages when communication is contextualized. Figure 1 plots the proportion of optimal languages under each objective as a function of context size. The red line shows that as the number of contexts increases, so does the probability that an optimal language $\ell^*_{cross}$ contains ambiguity under our CE objective. For comparison we also plot the proportion of speaker-optimal $\ell^*_{speaker}$ (blue line) and listener-optimal $\ell^*_{listener}$ (green line) languages that contain ambiguity. In line with Zipf's predictions, if languages are designed only to minimize speaker effort then optimal languages always contain ambiguity. If languages are designed to minimize listener effort then ambiguity is always avoided.\par

While our results indicate that ambiguity is an efficient property of contextualized language, these simulations assumed that agents could always disambiguate items in context. That is, in our four conditions both agents had perfect knowledge of the relevant conditional distributions ($P(M|C)$). This assumption may be too strong for describing much of day-to-day communication -- we seldom interact with others with perfect knowledge of the current context (or topic) at the start of a conversation. To explore how ambiguity may be \textit{used} efficiently in our framework, we next examine a case in which the listener has imperfect knowledge of context at the start of the conversation, but may infer it from the discourse history.\par

@@

We shift focus in Simulation 2 -- instead of examining the landscape of a space of languages under our CE objective, we consider a single language containing ambiguity, examining how that property is used over a discourse. If context is not informative with respect to meaning early in a discourse, but is at later positions, then an efficient speaking strategy would avoid ambiguous material early, saving it for when contextual information is disambiguating. We can consider “context” as analogous to a “topic” of conversation. As an example, imagine a scenario in which a reader is beginning a magazine article. While they may have some knowledge about the article’s topic (perhaps from the title), they may not have complete knowledge of its contents, including the persons or events involved. In this setting, using a low-cost, but ambiguous referring expression (say a pronoun like "he") early may lead to misunderstanding if context is not informative. But, if by a later position enough contextual information has accumulated, it may be efficient to use the ambiguous expression.\par

@@

We consider consider a single language $\ell$, which contains both ambiguous and unambiguous utterances. We assume ambiguous utterances are lower cost. Crucially, we do not assume that the listener knows the particular topic ($c_{current}$) of the conversation \textit{a priori}. Rather, that the listener has knowledge of the set of possible topics $C = \{c_1, \dots, c_k\}$, but does not know which one is currently being used by the speaker. Formally, this means the listener does not have access to the correct conditional distribution over meanings $P(M|C=c_{current})$ at the start of the discourse.\par

@@

We conduct $N=600$ simulations, generating discourses of length $|D|=30$ utterances, comparing three speaker models ($n=200$ each). We consider a single language $\ell$\footnote{See SI for the matrix notation of this langauge.} with $|U|=6$ and $|M|=4$ in which two of the utterances are ambiguous and lower cost than the unambiguous utterances. (Note that use of this particular language is not essential -- the results are broadly generalizable to languages that contain ambiguity, but exploring this space is computationally expensive.)\par

## Speaker agents
We consider three types of speaker models. Our \textit{Full pragmatics} agent, models a speaker who reasons about her listener and also has complete recall of the set of utterances in the discourse $D$. This speaker believes that the listener may not know the current topic $c_{current}$ at the start of the discrouse, but can infer it over the discourse. We compare two baseline models. The first, a \textit{Partial pragmatics} baseline describes a speaker who reasons about a listener, but assumes they have no access to discourse history. The second, a \textit{No pragmatics} speaker who does not consider a listener at all, but produces utterances according to the underlying langauge semantics ($\ell$) and topic probabilities ($p(M|C=c_{current}$) (see SI section 3).\par

@@

We are interested in when it is efficient to use low-cost, ambiguous material in this discourse setting. A speaker strategy that is mutually efficient for both agents should avoid ambiguity until sufficient contextual information has accumulated. We should expect this to be reflected in our \textit{Full pragmatics} model who reasons about the listener and discourse history. By contrast, a speaker-optimal model who does not consider the listener should greedily use ambiguous utterances (\textit{No Pragmatics} model), while a listener-optimal model should avoid ambiguity entirely (\textit{Partial pragmatics} model).\par

@@

Figure 2 (A) shows the empirical probability that a speaker uses an ambiguous utterance as a function of discourse position. The \textit{No pragmatic} baseline uses ambiguous utterances frequently and at a constant rate over the discourse and the \textit{Partial pragmatic} baseline avoids ambiguous utterances entirely. But, the \textit{Full pragmatic} model avoids ambiguous material only at the start of the discourse, employing it increasingly as the discourse proceeds. Panel (C) tracks our CE objective for each model over the discourse. Note that the objective decreases for all three models, primarily driven by the listener updating his belief about the actual topic ($P(C=c_{current}|D)$). However, the objective declines more quickly under the \textit{Full} and \textit{Partial} pragmatic speakers as listener agents are better able to infer the correct context. Additionally, the difference in CE between the \textit{Full}- and \textit{Partial pragmatic} models at the end of the discourse is driven by the reduction in speaker costs (see panel B). While speaker effort remains constant in both \textit{No pragmatic} and \textit{Partial pragmatic} baselines, effort declines in the \textit{Full pragmatic} model as she increasingly relies on ambiguous material later in the discourse.\par

@@

How do the competing pressures of speakers and listeners give rise to the distributional regularities found in natural language? Zipf [-@Zipf1949a] proposed that the asymmetry between speaker and listener costs gives rise to a range of properties at the level of the lexicon. We explored the interactions of rational pragmatic agents as a framework for understanding efficient language structure and use. We focused on an argument on the communicative function of ambiguity [@Piantadosi2011a], deriving a novel speaker-listener Cross-Entropy objective for measuring the efficiency linguistic systems from first principles of efficient language use. In a series of simulations we showed that optimal languages are more likely to contain ambiguous material when context is informative and that rational pragmatic agents will use ambiguous material efficiently.\par

A limitation of the current work is an analysis of exactly how this objective compares to existing measures. For example, previous work has described competing speaker-listener pressures in terms of a trade-off of simplicity and informativeness [@KempRegier2012a] or expressivity and compressibility [@SmithTamarizKirby2013a]. Future work should assess the degree to which we can derive the same properties as previous studies using our current framework. In line with this idea, a primary focus of the current work is to provide a case-study on the productivity of assessing functionalist theories in repeated reference games between rational pragmatic agents. We hope this framework can serve as a domain general tool to assess the range of functionalist theories examining efficient language-structure and use.\par
